%!TEX root = main.tex
\section{The CASTM API}
\label{sec:openedge}

\castm{}\footnote{Available at \url{https://github.com/PhDCristian/CASTM}}, can be seen as a deterministic C-like Domain Specific Language (DSL) whose design principle is \emph{explicit spatial-temporal control with structured abstraction}: the programmer specifies what executes on which PE at which cycle, but through high-level constructs instead of raw encodings.

\subsection{API Design}
\label{subsec:lang-overview}

A \castm{} program declares a \emph{target profile}, a \emph{build configuration}, and one or more \emph{kernel} definitions. Table~\ref{tab:openedge-constructs} organizes the constructs into three layers: spatial-temporal statements (instruction placement), structured abstractions (functions, loops, labels), and architecture-aware operators for modular arithmetic idioms.

\begin{table}[hbt]
\caption{\castm{} API constructs by abstraction layer.}
\label{tab:openedge-constructs}
\centering
\scriptsize
\setlength{\tabcolsep}{2pt}
\begin{tabular}{@{}llp{3.2cm}@{}}
\toprule
\textbf{Layer} & \textbf{Construct} & \textbf{Semantics} \\
\midrule
\multirow{7}{*}{\parbox{1.2cm}{Spatial-\\temporal\\statem.}}
 & \texttt{bundle \{\ldots\}} & Group instr.\ into one bundle \\
 & \texttt{@r,c: instr;} & Single-PE at (row, col) \\
 & \texttt{at @r..r,c..c:} & Range across PE sub-grid \\
 & \texttt{at all: instr;} & Broadcast to all PEs \\
 & \texttt{R0 = R1 + R2;} & C-like expr.\ $\to$ ALU op \\
 & \texttt{LWI/SWI R0, 40;} & Mem load/store (13-bit) \\
 & \texttt{BSFA R3,R0,R2;} & Branchless select \\
\midrule
\multirow{3}{*}{\parbox{1.2cm}{Struct.\\abstr.}}
 & \texttt{function f(a,b)} & Parametric macro \\
 & \texttt{for c in range(N)} & For loop \\
 & \texttt{goto label/R3;} & Direct/indirect jump \\
\midrule
\multirow{5}{*}{\parbox{1.2cm}{Arch.-\\aware\\oper.}}
 & \texttt{extract\_bytes} & Byte extraction \\
 & \texttt{carry\_chain} & Multi-limb carry prop. \\
 & \texttt{broadcast} & Wave to all PEs \\
 & \texttt{normalize} & Limb norm.\ + carry \\
 & \texttt{reduce} & Collective accumul. \\
\bottomrule
\end{tabular}
\end{table}

The \emph{spatial-temporal statements} layer provides direct control over instruction placement: every statement maps one-to-one to a CSV entry.
The \emph{structured abstractions} layer reduces repetition without scheduling opacity: \texttt{function} expands the function body at the call site by default (parametric macro expansion)  but can be compiled to shared subroutines under the \texttt{jump-reuse}  expansion mode, which emits each function body once and links calls through a register-based return convention.
The \emph{architecture-aware operators} layer encapsulates multi-instruction CGRA idioms for modular arithmetic, each expanding to a fixed, documented sequence.

To illustrate the contrast with manual CSV, consider broadcasting a memory load to all 16 PEs. In CSV, this requires 16 identical entries per cycle (each encoding opcode, mux selects, register indices, and immediate as 7 numeric fields): \fbox{\texttt{29,0,0,0,1,0,0}} repeated 16 times. In \castm{}, the same operation is \mbox{\texttt{bundle \{ at all: LWI R0, 0; \}}} --a single line that the compiler expands to 16 entries, eliminating an entire class of copy-paste errors--.

\subsection{Compilation Pipeline}
\label{subsec:pipeline}

The compiler performs a four-stage deterministic pipeline:
\textbf{Parse} (tokenize, syntax validate) $\to$ \textbf{Analyze} (semantic checks, function expansion, constant resolution) $\to$ \textbf{Lower} (operator expansion, loop, flat IR generation) $\to$ \textbf{Emit} (per-PE CSV encoding). Each stage is a pure transformation with no heuristic decisions. The compiler rejects ambiguous scheduling (e.g., two instructions on the same PE in the same cycle) with diagnostics reporting source location, PE coordinate, and cycle number.

\castm{} combines three properties no prior CGRA tool provides simultaneously: \emph{deterministic compilation with explicit placement} (cycle count predictable at authoring time), \emph{architecture-aware operators}
and \emph{transparent expansion} (every construct has a documented, predictable expansion to CSV).

\subsection{S-box Kernel Walkthrough}
\label{subsec:sbox-openedge}

Fig.~\ref{fig:sbox-pipeline} illustrates how our S-box kernel decomposes $x^7 \Mod p$ into four (modular) phases: 1) $x^2$ (\texttt{subrC} --squaring--); 2) $x^3 = x^2 \cdot x$ (\texttt{subrB} --general multiplication--); 3) $x^4 = (x^2)^2$ (\texttt{subrC}); and 4) $x^7 = x^4 \cdot x^3$ (\texttt{subrB}). Each phase performs a Barrett modular multiplication that consists of a base multiplication (\texttt{subrB}/\texttt{subrC}) followed by a Barret modular reduction (\texttt{subrA}).

\begin{figure}[t]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[
    block/.style={draw, thick, minimum height=0.45cm, font=\scriptsize\sffamily\bfseries, rounded corners=1pt, text=white, anchor=west, inner xsep=0pt, outer sep=0pt},
    lbl/.style={font=\normalsize\sffamily, cgragray},
]

% =========================================================
% 1. VARIABLES DE CONFIGURACIÓN GENERAL
% =========================================================
\def\totalw{8.0} % Ancho total físico en cm
\pgfmathsetmacro{\s}{\totalw/\MetricSboxKernelComputeCC} % Factor de escala

% Latencias en ciclos de reloj (CC)
\def\wA{39}
\def\wB{13}
\def\wC{8}
\def\gap{2} % Hueco de separación entre operaciones completas

% Alturas Y para organizar visualmente
\def\yBlock{0.25} 
\def\yLabel{0.55} 

% =========================================================
% 2. DIBUJO DEL EJE TEMPORAL
% =========================================================
\draw[thick, -{Stealth[length=5pt, width=4pt]}] (0,-0.1) -- (\totalw+0.5,-0.1);

\foreach \x in {0, 50, 100, 150, \MetricSboxKernelComputeCC} {
    \pgfmathsetmacro{\xpos}{\x*\s}
    \draw[thick] (\xpos,-0.1) -- (\xpos,-0.3);
    \node[lbl, below] at (\xpos,-0.3) {\x};
}
\node[lbl, right] at (\totalw+0.1, 0.15) {CC};

% =========================================================
% 3. MACRO DE GENERACIÓN DE ETAPAS DEL PIPELINE
% =========================================================
% Inicializamos la variable global para rastrear el ciclo actual
\def\currentCC{0}

% Comando: \drawSboxStage{Fórmula}{Texto 1}{Ancho 1}{Color 1}{Texto 2}{Ancho 2}{Color 2}
\newcommand{\drawSboxStage}[7]{
    % 1. Dibuja el primer bloque (op. auxiliar: ej. C o B)
    \node[block, fill=#4, minimum width=#3*\s cm] at (\currentCC*\s, \yBlock) {#2};
    
    % 2. Calcula el inicio del bloque principal
    \pgfmathparse{\currentCC + #3}
    \let\startMain\pgfmathresult
    
    % 3. Dibuja el bloque principal (reducción de Barrett: ej. A)
    \node[block, fill=#7, minimum width=#6*\s cm] at (\startMain*\s, \yBlock) {#5};
    
    % 4. Centra la fórmula matemáticamente sobre el bloque principal
    \pgfmathparse{(\startMain + #6/2)*\s}
    \node[font=\Large\sffamily\bfseries, above] at (\pgfmathresult, \yLabel) {#1};
    
    % 5. Actualiza la variable global para la siguiente etapa (suma el gap)
    \pgfmathparse{\startMain + #6 + \gap}
    \xdef\currentCC{\pgfmathresult} % \xdef fuerza la asignación global
}

% =========================================================
% 4. EJECUCIÓN DEL PIPELINE (Limpio y escalable)
% =========================================================
\drawSboxStage{$x^2$}{C}{\wC}{cgrared!80}{A}{\wA}{cgrablue!80}
\drawSboxStage{$x^2\!\cdot\!x$}{B}{\wB}{cgraorange!80}{A}{\wA}{cgrablue!80}
\drawSboxStage{$x^4$}{C}{\wC}{cgrared!80}{A}{\wA}{cgrablue!80}
\drawSboxStage{$x^7$}{B}{\wB}{cgraorange!80}{A}{\wA}{cgrablue!80}

\end{tikzpicture}%
}
\caption{Temporal breakdown of the S-box kernel (\cc{\MetricSboxKernelComputeCC}). Barrett modular reduction (\texttt{subrA}) consumes \pct{\MetricSboxKernelBarrettSharePct} of execution.}
\label{fig:sbox-pipeline}
\end{figure}

Fig.~\ref{fig:barrett-structure} shows the 6-steps Barret algorithm~\cite{barrett1986} and how it is decomposed into four algorithmic stages mapped to a specific CGRA hardware capability. Step 1 represents the base product (\texttt{subrB}/\texttt{subrC}) while steps 2-6 correspond to the Barret modular reduction (\texttt{subrA}). Step 1 is implemented as a parallel convolution: 32-bit operands decomposed into four 8-bit limbs, producing 16 partial products on the $4\!\times\!4$ PE mesh in our CGRA. Steps 2-3 represent the quotation estimation and are mapped as a toroidal accumulation via nearest-neighbor routing; Step 4, the modular subtraction, avoids expensive division via subtraction and AND masking; then steps 5-6 perform the final modular correction via branchless conditional BSFA (a single-cycle register select based on the ALU sign flag).

\input{figures/fig_barrett_structure_test}

\begin{figure}[hbt]
\begin{lstlisting}[caption={S-box kernel for $x^7 \Mod p$ (key excerpts). \texttt{RCR/RCL/RCT/RCB} denote \textit{Read Cell Right, Left, Top, Bottom}.}, label={lst:sbox-kernel}, breaklines=false, basicstyle=\ttfamily\fontsize{7}{8.4}\selectfont, columns=fullflexible, xleftmargin=2mm]
kernel "SBox" {
 // Main: set return addr, jump to subrC
 bundle { @0,1: goto subrC;
           @3,0: R3 = ZERO + ret1; }
 ret1: //(*@\textcolor{cgragreen!70!black}{$\,\triangleright$ triple merge}@*)
 bundle { @0,0: SWI R3, 720;
           @0,1: goto subrB;
           @3,0: R3 = ZERO + ret3; }
 ...
 subrC: // squaring accumulation (8 bndl)
 extract_bytes_col(R0, R1); //(*@\textcolor{cgragreen!70!black}{$\,\triangleright$ 4 PEs extract bytes}@*)
 extract_bytes_row(R0, R0);
 bundle { at all: R2 = R0 * R1; } //(*@\textcolor{cgragreen!70!black}{$\,\triangleright$ 16 products}@*)
 // Toroidal accumulation (4 bndl, RCR/RCT/RCB)
 bundle { @3,0: goto subrA; }

 subrA: // Barrett reduction (39 bndl)
 bundle { @0,0: R1 = RCR;
           @0,1: R0 = RCB;
           @0,2: R1 = RCB; }
 // Quotient estimation: 12 PEs
 bundle { at @0..2,0..3: R2 = R0 * R1; }
 // ... carry prop., normalization
 // Branchless correction:
 bundle { @0,0: R2 = R0 - RCL; }
 bundle { @0,0: BSFA R3, R0, R2,S ELF;}//(*@\textcolor{cgragreen!70!black}{$\,\triangleright$ if r$\geq$p: r-p}@*)
 bundle { @3,0: goto R3; } //(*@\textcolor{cgragreen!70!black}{$\,\triangleright$ return}@*)
}
\end{lstlisting}
\end{figure}

Listing~\ref{lst:sbox-kernel} shows key excerpts of the S-box kernel.
The kernel structure reveals several design decisions enabled by \castm{}:

\textbf{Jump-reuse subroutines}. All calls to subroutines are performed via register-indirect jumps. The Barrett modular reduction phase (\texttt{subrA}, \MetricSboxKernelSubrABundles~bundles) is shared across all four modular operations through a simple calling convention: the caller stores a return-address label in R3 (\mbox{\texttt{R3 = ZERO + retN}}, line 3, line 6), and the subroutine returns with \mbox{\texttt{goto R3}} (line 25). This avoids duplicating \MetricSboxKernelSubrABundles~bundles $\times$ \MetricSboxKernelSubrACalls~calls = \MetricSboxKernelSharedBundlesSaved~wasted bundles, reducing total kernel size from ${\sim}\MetricSboxKernelApproxPreReuseBundles$ to \MetricSboxKernelResidentBundles~bundles (see Section \ref{subsec:sbox-comparison}). In \castm{}, this pattern is expressed naturally with labels and \texttt{goto}; in manual CSV, the developer must compute absolute addresses for every jump target.

\textbf{Triple instruction merging.} At label \texttt{ret1}, three independent operations execute on three different PEs in the same cycle (lines 5-6): \texttt{SWI} stores the intermediate result, \texttt{goto} jumps to the next subroutine, and \texttt{SADD} prepares the return address. The \texttt{bundle\{\}} statement makes this spatial parallelism explicit and verifiable; in CSV, it is easy to accidentally place two instructions on the same PE, causing a silent overwrite.

\textbf{Systolic convolution.} The \texttt{subrC} uses the $4\!\times\!4$ mesh as a spatial multiplier (lines 10-14): \texttt{extract\_bytes} operators decompose operands into byte-size limbs distributed across columns and rows, then \mbox{\texttt{at all: R2 = R0 * R1}} computes all 16 partial products simultaneously. Toroidal accumulation converges results in 4~cycles via neighbor reads (RCR, RCT, RCB). All 16 PEs compute their partial product simultaneously; for squaring, the symmetric pairs are folded via in-place doubling during the subsequent accumulation phase, reducing it from 13 to 8 bundles compared to \texttt{subrB} that accumulates all the partial results explicitly.

\textbf{Branchless correction.} The final BSFA instruction (line 24) selects between $r$ and $r - p$ based on the sign flag, completing modular reduction in a single cycle without branching---avoiding control divergence across the PE mesh.

The jump-reuse feature imposes a residency constraint: the full \MetricSboxKernelResidentBundles-bundles of the kernel must remain simultaneously resident in the Configuration Register File --CRF-- so that register-indirect control flow can execute without fragmentation. When the kernel no longer fits, it must be partitioned into multiple dispatchable segments, sacrificing subroutine sharing and introducing reload overhead. Section~\ref{subsec:creg-analysis} quantifies this S-box-specific CRF-pressure trade-off and shows why it becomes a first-order architectural constraint for the full Poseidon2 evaluation.

\subsection{Spatial Mapping Methodology}
\label{subsec:spatial-mapping}

The S-box mapping described above relies on a set of design patterns that are not specific to Poseidon2 or BabyBear, but apply to any modular-arithmetic kernel on a $N{\times}M$ toroidal CGRA. We describe them here as reusable principles.

\subsubsection{Operand Decomposition across the PE Mesh}
\label{subsubsec:decomposition}

When an arithmetic operation produces intermediates wider than the PE datapath ($w$~bits), the standard approach on a sequential processor is to serialize limb-by-limb. On a CGRA, the 2D mesh can be exploited spatially: each operand is decomposed into $N$ limbs of $w/N$~bits, one per PE along a column (or row), so that all limbs are available simultaneously without memory contention.

In our case, two 31-bit operands $a$ and $b$ are decomposed into four 8-bit limbs distributed along columns and rows, respectively. The $4\!\times\!4$ grid then computes all 16 partial products $a_i \cdot b_j$ in a single bundle---the mesh \emph{is} the multiplier array. The anti-diagonals ($k = i+j$) naturally group products by significance (base-$2^8$ weight), so the subsequent accumulation reduces to summing along these diagonals via nearest-neighbor links.

This pattern generalizes to any field width: for a 64-bit field (e.g., Goldilocks) one would use 16-bit limbs on the same $4\!\times\!4$ mesh, or 8-bit limbs on a larger array. The mesh dimensions naturally match the number of limbs, making the CGRA topology a first-class design parameter rather than an afterthought. For squaring ($a = b$), the symmetry $a_i \cdot a_j = a_j \cdot a_i$ allows the accumulation phase to fold off-diagonal pairs via in-place doubling, reducing accumulation from 13 to 8~bundles compared to general multiplication.

\subsubsection{Toroidal Carry Propagation}
\label{subsubsec:carry-chain}

After computing partial products, carries must propagate from less-significant to more-significant limbs. On a conventional processor this is a sequential chain; on the CGRA, it becomes a spatial wavefront through the nearest-neighbor interconnect. Carries propagate left-to-right along row~0 via \texttt{RCL} (Read Column Left) in a repeating three-instruction pattern per hop: \textit{extract} carry (\texttt{R3 = R0 {>}{>} 16}), \textit{mask} the current limb (\texttt{R0 = R0 \& 0xFFFF}), and \textit{inject} the carry into the next column (\texttt{R0 = R0 + RCL}). Vertical aggregation from rows below uses \texttt{RCB}/\texttt{RCT} (torus links), and the torus wrap bounds the maximum propagation distance to $N$~hops regardless of mesh size.

This spatial carry pattern replaces what would be a multi-cycle sequential loop on the host processor with a fixed-latency wavefront whose depth equals the mesh diameter. The principle is directly applicable to any multi-limb operation (Montgomery multiplication, Karatsuba decomposition, NTT butterfly networks) on any toroidal CGRA, regardless of the specific prime or field size.

\subsubsection{Subroutine Sharing on a Stack-less Architecture}
\label{subsubsec:subroutine-sharing}

CGRAs typically lack a call stack: the program counter advances through the CRF linearly or via direct jumps. Mapping modular exponentiation na\"ively would inline every Barrett reduction, producing a kernel whose size grows linearly with the number of modular operations---$O(n \cdot B)$ bundles for $n$ reductions of $B$ bundles each. For $x^7 \Mod p$ with $n{=}4$ and $B{=}39$, this alone would exceed 150~bundles before accounting for the multiplications themselves.

Our approach introduces a \emph{register-indirect return convention}: before calling a subroutine, the caller writes the return-address label into a designated register (R3) at a fixed PE (PE[3,0]). The subroutine concludes with \texttt{goto R3}, transferring control back to the caller. This reduces kernel size from $O(n \cdot B)$ to $O(B{+}n)$---the subroutine body appears once, plus a small dispatch stub per call.

The convention also enables \emph{fall-through chaining}: when two subroutines are logically sequential (e.g., multiplication followed by reduction), placing them contiguously in instruction memory eliminates the jump entirely---execution falls through from the last bundle of one into the first of the next. This asymmetric layout (squaring uses an explicit jump to reduction; multiplication falls through) saves one cycle per multiplication call.

\subsubsection{Multi-PE Bundle Merging}
\label{subsubsec:merging}

A CGRA bundle activates all PEs simultaneously, but not every PE needs to perform useful work in every cycle. When independent operations target different PEs, they can be \emph{merged} into the same bundle without conflict. The S-box kernel exploits this systematically: at return points, three operations execute in a single cycle on three distinct PEs---a memory store (\texttt{SWI} on PE[0,0]), a subroutine jump (\texttt{goto} on PE[0,1]), and a return-address setup (\texttt{R3 = ZERO + retN} on PE[3,0]).

This pattern is especially effective at subroutine boundaries, where control-flow overhead (jumps, address setup) would otherwise consume dedicated cycles. The general principle is that any CGRA kernel should be audited for \emph{bundle utilization gaps}---cycles where some PEs are idle---and those gaps should be filled with logically independent operations from adjacent phases. In our case, this transforms what would be three sequential overhead cycles into zero additional latency.
